---
title: "Data summary and processing details"
author: "Fangwu Wang"
date: '2017-02-14'
output: 
 html_document:
  toc: yes
  keep_md: true
---
All of the data is from BluePrint Epigenome -- 
Data portal link: http://www.medical-epigenomics.org/papers/BLUEPRINT_methylomes/#gbrowser

Reference: Farlik M, Halbritter F, Müller F, Choudry FA et al. DNA Methylation Dynamics of Human Hematopoietic Stem Cell Differentiation. Cell Stem Cell 2016 Dec 1;19(6):808-822. PMID: 27867036

#Description of DNA methylation data: 

The DNA methylation profiles are available through GEO (Gene Expression Omnibus) in the alignment format. Due to large sizes, we can not deposit the data into GitHub team repo.  

Samples: The paper generated small pools (50 cells, 100 cells) of DNA methylomes on each sample to increase the coverage and accuracy of detection. We chose three samples (biological replicates) with multiple small pool profiles for each population. Younger ages are prefered and the number and mean coverage of sites are taken into consideration for quality check.  
Quality report is available at RnBead website: http://www.medical-epigenomics.org/papers/BLUEPRINT_methylomes/data/rnbeads_reports/prog_v08_um_nsc/index.html

Processing: The available DNA methylation data has been aligned and not quantitated. The procedure of raw data processing is from the supplementary methods of the reference paper.  
1). Bisulfite sequencing was done by the Biomedical Sequencing Facility at CeMM using the 2x75bp paired-end setup on the Illumina HiSeq 3000/4000 platform.  
2). Sequencing adapter fragments were trimmed using Trimmomatic v0.32. The trimmed reads were aligned with Bismark v0.12.2 (Krueger and Andrews, 2011) with the following parameters: --minins 0 --maxins 6000 --bowtie2, which uses Bowtie2 v2.2.4 (Langmead and Salzberg, 2012) for read alignment.  
3). The GRCh38 assembly of the human reference genome was used throughout the study, in a version for sequence alignment obtained from NCBI.  
4). Duplicate reads were removed as potential PCR artefacts, and reads with a bisulfite conversion rate below 90% or with fewer than three cytosines outside a CpG context (required to confidently assess bisulfite conversion rate) were removed as potential post-bisulfite contamination. 


Summary of DNA methylation data:
```{r echo = FALSE}
a<-read.delim("/Users/fangwu/Documents/540/data/DNA.txt")
library(knitr)
kable(a)
```


#Description of RNA-seq data: 

##**RNA-seq for normal samples (matched to DNA methylation)**
The RNA-seq data was from the same study and obtained from GEO (GSE87195). The only one data file "GSE87195_rnaseq_ensT_all.csv" was deposited under "Data" folder of our team repo. The raw data is not directly available for RNA-seq. The uploaded file has been almost fully processed and the output is the format after differential expression analysis using the BitSeq software. The data is quantified at transcript level with ENST IDs.

Samples: This single file contains all seven populations of interest corresponding to the DNA methylation data with the same tissue origin.  

Processing:
1). After RNA-seq on Illumina HiSeq 2000 platform, sequencing adapter fragments were trimmed using Trimmomatic v0.32.  
2). The trimmed reads were aligned to the cDNA reference transcriptome (GRCh38 cDNA sequences from Ensembl) using Bowtie v1.1.1 and the following parameters: -q -p 6 -a -m 100 --minins 0 --maxins 5000 --fr --sam --chunkmbs 200.  
3). Duplicate reads were removed, and transcript levels were quantified with BitSeq v1.12.0. Transcript-level expression were estimated.  

Summary of DNA methylation data:
```{r echo = FALSE}
b<-read.delim("/Users/fangwu/Documents/540/data/RNA.txt")
library(knitr)
kable(b)
```

##**RNA-seq for leukemia samples**

Two types of leukemia samples--Acute myeloid leukemia and chromic lymphocytic leukemia--were derived from BluePrint Epigenome consortium: http://dcc.blueprint-epigenome.eu/#/experiments

Processing:
The GRAPE 2(https://github.com/guigolab/grape-nf) pipeline has been used to process the data. The used profile is STAR-RSEM, adapted from the ENCODE Long RNA-Seq pipeline (https://github.com/ENCODE-DCC/long-rna-seq-pipeline).  
1). Mapping: The mapping step is performed using the [STAR] spliced aligner to map the RNA-seq reads to the references, which can be found in the Blueprint references FTP (ftp://ftp.ebi.ac.uk/pub/databases/blueprint/reference/20150407_reference_files).  
2). Signal: In this step the STAR (https://github.com/alexdobin/STAR/) spliced aligner is used to produce signal files. The procedure is differentiated for strand-specific and non-strand-specific samples. For strand-specific data the signal from the two strands is separated. For non-strand-specific data the signal from the two strands is collapsed.   
3). Contigs: In this step BEDtools (http://bedtools.readthedocs.org/en/latest) and the contigsNew Python script (https://github.com/guigolab/grape-nf/blob/develop/bin/contigsNew.py) are used to produce a BED file with the contig information. 
4). Quantification: In this step RSEM (http://deweylab.biostat.wisc.edu/rsem/) is used to provide transcript and gene quantifications. A specific SAMtools version (https://github.com/emi80/samtools/tree/1.2-rglab-CRG) is used to systematically sort the reads as required by RSEM (http://deweylab.biostat.wisc.edu/rsem/) .  
Output files ( RSEM output format (http://deweylab.biostat.wisc.edu/rsem/rsem-calculate-expression.html#output)):  
${prefix}.genes.results
${prefix}.isoforms.results

Summary of DNA methylation data:
```{r echo = FALSE}
c<-read.delim("/Users/fangwu/Documents/540/data/RNA-L.txt")
library(knitr)
kable(c)
```





